---
title: "Chapter 1 - Importing and Visualizing Functional Data"
author: "Vanilton Paulo"
date: "2025-08-15"
output:
  html_document: default
---

# COVID-19 US Mortality Data

#### Weekly all-cause mortality data

We assemble exploratory visualizations for all-cause and COVID-19 weekly mortality data in the US. The data can be accessed via the `refund` package in R. The all-cause mortality data is the weekly mortality that covers the period from the week ending January 14, 2017, through the week ending April 10, 2021.

The COVID-19 mortality data contain weekly mortality, with COVID-19 listed as the underlying cause of death. The data span from the week ending January 4, 2020, to the week ending April 17, 2021, and are available via the `refund` package in R.

We use the `tidyfun` package to visualize U.S. mortality trends with a focus on the year 2020. The analysis addresses two objectives:

1.  To visualize all-cause total and excess mortality(mortality in a week in 2020 minus mortality in the corresponding week in 2019).

2.  To visualize total COVID-19 mortality as a function of time.

```{r setup,echo = TRUE, message = FALSE}
# ─ Packages ───────────────────
# List of required packages
packages <- c(
  "refund",
  "tidyverse",
  "tidyfun"
)

# Install missing packages
install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# Install missing packages
invisible(lapply(packages, install_if_missing))



# Load packages
library(refund)      
library(tidyverse)   
library(tidyfun)     

# For reproducible results
set.seed(53615)
```

```{r load-data, echo = TRUE, results = 'hide'}
# ──── Overview ───────────────────
data("COVID19", package = "refund") 
names(COVID19)
##  [1] "US_weekly_mort"                      "US_weekly_mort_dates"               
##  [3] "US_weekly_mort_CV19"                 "US_weekly_mort_CV19_dates"          
##  [5] "US_weekly_excess_mort_2020"          "US_weekly_excess_mort_2020_dates"   
##  [7] "US_states_names"                     "US_states_population"               
##  [9] "States_excess_mortality"             "States_excess_mortality_per_million"
## [11] "States_CV19_mortality"               "States_CV19_mortality_per_million"
```

```{r helpers,echo = TRUE}
# ───────────── helpers ───────────────────
# Converts calendar dates to numeric values 
# The tfd requires numeric values and not as dates values.
num_grid <- function(dates, ref = min(dates)) as.numeric(dates - ref)
reference_date <- as.Date("2020-01-01")
```

#### Weekly all-cause mortality in the US

```{r data, include = TRUE}
# Obtain weekly all-cause excess mortality between 2017 to end of 2020 
counts_state <- COVID19$US_weekly_mort
date_state <- COVID19$US_weekly_mort_dates
```

Visualization of  all-cause mortality between 2017 to end of 2020

```{r data-transformation, include = TRUE}
# ─── Data Transformation ─────────────
nat_weekly <- tibble(
  date = date_state,
# The weekly all-cause death counts  
# Divide by 1000 to indicate numbers in thousands
  deaths = counts_state / 1000       
)


# Shaded regions highlighting 2019 (blue) and 2020 (red)
shade <- nat_weekly %>%
  filter(year(date) %in% c(2019, 2020)) %>%
  mutate(
    year = year(date),
    day_num = num_grid(date, ref = min(nat_weekly$date))
  ) %>%
  group_by(year) %>%
  summarise(
# For each year, get the min and max week-index
    xmin = min(day_num),
    xmax = max(day_num),
    .groups = "drop"
  ) %>%
  mutate(
    band = as.character(year),
    colour = c("blue", "red")
  )

# X-axis year labels at first week of each year
year_ticks <- nat_weekly %>%
# We add two new columns
  mutate(
    year = year(date),
    day_num = num_grid(date)
  ) %>%
  group_by(year) %>%
# Within each year, we take the row with the smallest date for the plot
  slice_min(date, n = 1) %>%
  ungroup()
```


```{r , include=TRUE}
# ────────── Turn into tfd ─────────────
nat_weekly_tfd <- tfd(
# We reshape the deaths into a 1-row matrix so it’s  one curve through 2017 to 2020.
  matrix(nat_weekly$deaths, nrow = 1),
  arg = num_grid(nat_weekly$date)
)
# tfd[1] on (0,1442) based on 207 evaluations each
# interpolation by tf_approx_linear 
# [1]: ( 0,61);( 7,59);(14,58); ..
```


```{r weekly-all-cause-plot, include = TRUE}
# ─────────────────── Plot ─────────────

ggplot() +
  geom_rect(
    data = shade,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = colour),
    inherit.aes = FALSE, alpha = 0.15
  ) +
  scale_fill_identity() +
  geom_meatballs(
    data = tibble(y = nat_weekly_tfd),
    aes(y = y),
    colour = "steelblue", size = 2, alpha = 0.6
  ) +
  scale_x_continuous(
    breaks = year_ticks$day_num,
    labels = year_ticks$year,
    expand = c(0, 0)
  ) +
  labs(
    x = "Weeks starting in January 2017",
    y = "Weekly all-cause deaths in the US (thousands)"
  ) +
  theme_classic() 
```

FIGURE 1.1: The weekly death counts in the United States from 14 of January 2017 to 26 of December 2020. Each point corresponds to a single week, with deaths reported in thousands. For instance, in the week of 5 May 2018 , there were 53,874 deaths nationwide.

#### Weekly US COVID-19 mortality data

```{r load-data-weeklycv19, include = TRUE}
# ─── Data ──────────────────
# Obtain covid-19 associated deaths in the US
cv19_deaths <- COVID19$US_weekly_mort_CV19
week_diff <- COVID19$US_weekly_excess_mort_2020
current_date <- COVID19$US_weekly_excess_mort_2020_dates
```

The COVID-19 mortality data contain weekly mortality, with COVID-19 listed as the underlying cause of death.
```{r data-transformation-weeklycv19, echo = TRUE}
# ─── Turn into tfd ──────────────────

death_tf <- tfd(
  rbind(week_diff, cv19_deaths),
  arg = num_grid(current_date)
)
# tfd[2] on (0,357) based on 52 evaluations each
# interpolation by tf_approx_linear 
# week_diff: ( 0,1698);( 7,2222);(14,1011); ...
# cv19_deaths: ( 0,   0);( 7,   0);(14,   3); ...

# ── Prepare data for plotting ──────────────────

excess_tibble <- tibble(
  death_type = c("All-cause excess", "COVID-19"),
  curve = death_tf
)
# A tibble: 2 × 2
#   death_type                                         curve
#   <chr>                                          <tfd_reg>
# 1 All-cause excess [1]: ( 0,1698);( 7,2222);(14,1011); ...
# 2 COVID-19         [2]: ( 0,   0);( 7,   0);(14,   3); ...


# ──────────────── Plot ────────────────
ggplot(excess_tibble, aes(y = curve, colour = death_type)) +
  geom_meatballs(size = 3, alpha = 0.6) +
  scale_colour_manual(values = c("blue", "red")) +
  scale_x_continuous(
    name = "Weeks starting January 2020",
    breaks = as.numeric(seq(reference_date, as.Date("2021-01-01"), by = "3 months") - reference_date),
    labels = format(seq(reference_date, as.Date("2021-01-01"), by = "3 months"), "%b %Y")
  ) +
  labs(
    x = "Weeks starting in January 2020",
    y = "All-cause excess and COVID-19 deaths in the US",
  color = "Type of death"
  ) +
  theme_classic() 
```

FIGURE 1.2: The total weekly U.S. mortality caused by COVID-19 alongside estimated excess deaths.The blue points represent excess mortality, calculated as the difference between the total deaths in a given week of 2020 and the corresponding week in 2019. The red points indicate the number of deaths each week directly caused by COVID-19.

Total excess deaths before the week ending on 26 december 2020.

```{r total-excess-calculation, include = TRUE}
cumsum(death_tf$week_diff)[52]
```

COVID-19 deaths before the week ending on on 26 december 2020.

```{r cv-calculation, include = TRUE}
cumsum(death_tf$cv19_deaths)[52]
```

There are 35.9% more excess deaths than COVID-19 deaths before 26 December 2020.

#### Plot of all-cause cumulative excess data

Plot the all-cause US excess mortality data for year 2020.

```{r load-data-all-cause, include = TRUE}
# ── Data ───────────────────
new_states <- COVID19$US_states_names
state_population <- COVID19$US_states_population
states_excess <- COVID19$States_excess_mortality
```

We highlight the following states: New Jersey (green), Louisiana (red), California (plum), Maryland (blue), and Texas (salmon).
```{r data-transformation-all-cause, include = TRUE}
# ── Data Transformation ───────────────────

# Normalize to rate per 1 million people
rate_mat <- sweep(states_excess, 1, state_population / 1e6, "/")
# Row by row , and replace each row with its cumulative sum.
# We transpose it back so that rows = states, columns = weeks again.
cum_mat <- t(apply(rate_mat, 1, cumsum))

# ────────────── Turn into tfd ──────────────────

#Thus tfd object, a data structure where each functional curve represents a distinct state.
state_tf <- tfd(cum_mat, arg = num_grid(current_date))
# tfd[52] on (0,357) based on 52 evaluations each
# interpolation by tf_approx_linear 
# [1]: ( 0, 0.8);( 7, 8.3);(14,-6.9); ...
# [2]: ( 0,  -5);( 7,  19);(14,  26); ...
# [3]: ( 0,  -7);( 7,  -1);(14,   5); ...
# [4]: ( 0,  -7);( 7,  21);(14,  20); ...
# [5]: ( 0,   5);( 7,   7);(14,  12); ...
#     [....]   (47 not shown)


# ── Prepare data for plotting ──────────────────

state_df <- tibble(
  state = new_states,
  curve = state_tf
)
# A tibble: 52 × 2
#    state                                                   curve
#    <chr>                                               <tfd_reg>
#  1 Alabama               [1]: ( 0, 0.8);( 7, 8.3);(14,-6.9); ...
#  2 Alaska                [2]: ( 0,  -5);( 7,  19);(14,  26); ...
#  3 Arizona               [3]: ( 0,  -7);( 7,  -1);(14,   5); ...
#  4 Arkansas              [4]: ( 0,  -7);( 7,  21);(14,  20); ...
#  5 California            [5]: ( 0,   5);( 7,   7);(14,  12); ...
#  6 Colorado              [6]: ( 0,   3);( 7,  10);(14,  14); ...
#  7 Connecticut           [7]: ( 0, 0.3);( 7, 1.1);(14, 4.2); ...
#  8 Delaware              [8]: ( 0,  14);( 7,  13);(14,   4); ...
#  9 District of Columbia  [9]: ( 0,   3);( 7,  11);(14,  60); ...
# 10 Florida              [10]: ( 0,   6);( 7,  15);(14,  24); ...
# ℹ 42 more rows

# ── Plot ────────────────

#Emphasized states
emph_states <- c("New Jersey", "Louisiana", "California", "Maryland", "Texas")
emph_cols <- c("darkseagreen3", "red", "plum3", "deepskyblue4", "salmon")

ggplot() +
  geom_spaghetti(
    data = state_df,
    aes(y = curve),
    colour = "grey20", alpha = .15
  ) +
  geom_spaghetti(
    data   = filter(state_df, state %in% emph_states),
    aes(y = curve, colour = state), alpha = 5
  ) +
  scale_colour_manual(values = setNames(emph_cols, emph_states)) +
  scale_x_continuous(
    name = "Weeks starting January 2020",
    breaks = as.numeric(seq(reference_date, as.Date("2021-01-01"), by = "3 months") - reference_date),
    labels = format(seq(reference_date, as.Date("2021-01-01"), by = "3 months"), "%b %Y")
  ) +
  coord_cartesian(ylim = c(-50, 2500)) +
  labs(
    x = "Weeks starting January 2020",
    y = "US states cumulative excess deaths per million"
  ) +
  theme_classic() +
  theme(legend.position = "right")
```

FIGURE 1.3: Each U.S. state,the lines track the cumulative weekly all-cause excess mortality per million for each U.S. state. Five states receive special focus and are distinguished by color: New Jersey (green), Louisiana (red), California (plum),Maryland (blue) and Texas (salmon).All remaining states are in grey including Puerto Rico and District of Colombia.

#### Plot of COVID-19 cumulative data

```{r load-data-cum, include = TRUE}
# ── Data ────────────
states_CV19_mort <- COVID19$States_CV19_mortality
```

```{r data-transformation-cum, include = TRUE}
# ── Data Transformation ───────────────────
# Each row of the mortality data is divided by the corresponding state's population (in millions)
per_mil_mat <- sweep(states_CV19_mort, 1, state_population / 1e6, "/")
# Cumulative sum of deaths per million over time for each state
cum_covid <- t(apply(per_mil_mat, 1, function(x) cumsum(replace_na(x, 0))))

# ────────────── Turn into tfd ──────────────────

covid_tf <- tfd(cum_covid, arg = num_grid(current_date))
# tfd[52] on (0,357) based on 52 evaluations each
# interpolation by tf_approx_linear 
# [1]: ( 0,0);( 7,0);(14,0); ...
# [2]: ( 0,0);( 7,0);(14,0); ...
# [3]: ( 0,0);( 7,0);(14,0); ...
# [4]: ( 0,0);( 7,0);(14,0); ...
# [5]: ( 0,0);( 7,0);(14,0); ...
#     [....]   (47 not shown)

# ── Data Transformation ───────────────────
covid_df <- tibble(
  state = new_states,
  curve = covid_tf
)
# A tibble: 52 × 2
#    state                                          curve
#    <chr>                                      <tfd_reg>
#  1 Alabama               [1]: ( 0,0);( 7,0);(14,0); ...
#  2 Alaska                [2]: ( 0,0);( 7,0);(14,0); ...
#  3 Arizona               [3]: ( 0,0);( 7,0);(14,0); ...
#  4 Arkansas              [4]: ( 0,0);( 7,0);(14,0); ...
#  5 California            [5]: ( 0,0);( 7,0);(14,0); ...
#  6 Colorado              [6]: ( 0,0);( 7,0);(14,0); ...
#  7 Connecticut           [7]: ( 0,0);( 7,0);(14,0); ...
#  8 Delaware              [8]: ( 0,0);( 7,0);(14,0); ...
#  9 District of Columbia  [9]: ( 0,0);( 7,0);(14,0); ...
# 10 Florida              [10]: ( 0,0);( 7,0);(14,0); ...
# ℹ 42 more rows

# ─────────── Plot ────────────────
#Plot of COVID-19 cumulative data

ggplot() +
  geom_spaghetti(
    data = covid_df,
    aes(y = curve),
    colour = "black", alpha = 0.1
  ) +
  # highlighted states
  geom_spaghetti(
    data   = filter(covid_df, state %in% emph_states),
    aes(y = curve, colour = state), alpha = 2
  ) +
  scale_colour_manual(values = setNames(emph_cols, emph_states)) +
  scale_x_continuous(
    name = "Weeks starting January 2020",
    breaks = as.numeric(seq(reference_date, as.Date("2021-01-01"), by = "3 months") - reference_date),
    labels = format(seq(reference_date, as.Date("2021-01-01"), by = "3 months"), "%b %Y")
  ) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(
    x = "Weeks starting January 2020",
    y = "Cumulative COVID-19 deaths per million (US states)",
    col = NULL
  ) +
  theme_classic() +
  theme(legend.position = "right")
```

FIGURE 1.4: Each U.S. state,the lines track the cumulative COVID-19 mortality for each U.S. state. Five states receive special focus and are distinguished by color: New Jersey (green), Louisiana (red), California (plum),Maryland (blue) and Texas (salmon).All remaining states are in grey including Puerto Rico and District of Colombia.
